---
title: "Analysis for Titz: Are psychologists constructing their scales in the right way..."
author: "Johannes Titz"
date: "04/27/2023"
#original date: "2/23/2022"
output:
  html_document:
    toc: true
---

This analysis accompanies the paper: Titz, J. Are psychologists constructing their scales in the right way? Insights from studying perfect unidimensionality.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, cache = F,
                      dev = c("cairo_pdf", "png"),
                      fig.path = "plots",
                      dev.args = list(antialias = "subpixel"),
                      dpi = 300,
                      echo = TRUE,
                      fig.height = 4.365, fig.width = 4.85,
                      options(width = 82))
```

# Part I: Theory

## Prepare

I use librarian to manage dependencies, so please install it if you do not have it yet (uncomment the first line).

```{r prepare}
# install.packages("librarian")
library(librarian)
shelf(partitions, pbapply, tidyverse, psych, johannes-titz/zysno, xtable, Matrix, plot.matrix)
# use multiple cores
cores <- round(parallel::detectCores() * 0.8)
# create vectors from phi matrix
vec_from_tbl <- function(a, b, c, d) {
  mtrx <- matrix(c(a, b, c, d), ncol = 2)
  v1 <- rep(c(0, 1), colSums(mtrx))
  v2 <- rep(c(0, 1), rowSums(mtrx))
  cbind(v1, v2)
}
```

## Generate Data

```{r generate, cache = F}
n <- 200
# all combos, even the ones where phi is NA
grid <- as.matrix(t(compositions(n = n, m = 4)))
colnames(grid) <- paste0("Var", 1:4)
# rowwise does not work with the following code, I do not know why
# multiple cores seem to give no advantage, so is not used (argument cl)
# we cache this manually as it takes some time
grid <- xfun::cache_rds({
  grid <- as_tibble(grid) %>%
    rowwise() %>%
    mutate(errors = Var3,
           expected_errors = (Var1 + Var3) * (Var3 + Var4) / (Var1+ Var2+ Var3+ Var4),
           h = 1 - errors / expected_errors)
  phi <- pbapply(grid[, 1:4], 1, function(x) phi(x, digits = 4))
  cbind(grid, phi)
}, file = "sim1", rerun = F)
grid <- as_tibble(grid)
```

```{r filter}
grid_unidim <-  grid %>%
  filter(Var3 == 0) # only unidimensional
grid_row <- grid %>%
  filter(phi >= 0, Var2 >= Var3) # error cell is b
```

## Plot

```{r phihs}
p <- ggplot(grid_row, aes(phi, h)) + 
  #geom_point(alpha = 0.05) +
  geom_hex(bins = 100, show.legend = FALSE) + 
  theme_classic() +
  scale_x_continuous("Phi", breaks = seq(0, 1, 0.1)) + 
  scale_y_continuous("H", breaks = seq(0, 1, 0.1) ) + 
  coord_fixed()
p
pdf("plots/phiH.pdf", width = 5, height = 5)
p
dev.off()
```

## FA model

It is not hard to create unidimensional model that results in a perfect bifactor solution with factor analysis (for Pearson Correlation):

```{r}
m1 <- vec_from_tbl(10, 80, 0, 10)
m2 <- vec_from_tbl(11, 78, 0, 11)
m3 <- vec_from_tbl(15, 70, 0, 15)

d <- cbind(m1, m2, m3[, -2])
# check that it is really unidimensional
zysnotize(d)

ncol <- ncol(d)
colnames(d) <- paste("Item", seq(ncol(d)))
f1 <- factanal(d, 1)
f2 <- factanal(d, 2)

f1$loadings
f2$loadings
print(xtable::xtable(cbind(f1$loadings, f2$loadings), label = "tab:fa1b", 
                     caption = "Cross tables of five unidimensional items"),
      file = "tables/fa1b.tex", booktabs = TRUE)

combs <- Map(function(x, y) table(d[, x], d[, y]),
             rep(1:ncol, each = ncol), rep(1:ncol, ncol))
lower <- seq(1, 21, 5)
upper <- seq(5, 25, 5)
res <- Map(function(x, y) Reduce(cbind, combs[x:y]),
    lower, upper)
res2 <- Reduce(rbind, res)
colnames(res2) <- paste0("$", rep(1:5, each = 2), "_", c(0, 1), "$")
rownames(res2) <- paste0("$", rep(1:5, each = 2), "_", c(0, 1), "$")

res2
print(xtable::xtable(res2, label = "tab:fa1",
                     caption = "Example of factor analysis for five unidimensional items"),
      booktabs = TRUE,
      file = "tables/fa1.tex",
      sanitize.text.function = function(x) {x})
```

# Part II: Empirical Example

Load data

```{r}
d <- read.csv2("data.csv")
```

Find order, calculate socre, show df

```{r}
o <- order(colMeans(d[,-1]), decreasing = TRUE)
d <- d[, c(1, o+1)]
d$score <- rowSums(d[, -1])
df <- as.data.frame(d)
df[order(df$score), ]
```

### Violations

```{r}
is_monotonic <- function(x) {
  all(diff(as.numeric(x)) <= 0)
}

df$monotonic <- apply(df[, 2:6], 1, is_monotonic)
dg <- df[df$monotonic == FALSE, ]

dh <- dg[order(dg$score), ]
dh
print(
      xtable::xtable(dh[, 2:7], caption = "Participant answers that violate unidimensionality", 
                     label = "tab:viol", digits = 0),
      booktabs = TRUE,
      file = "tables/viol.tex"
)
```

### unidim analysis

```{r}
lv <- loevenize(as.matrix(d[, 2:6]))
lv
```

### FA

```{r}

fa1 <- factanal(d[,c(-1, -7)], 1)
fa2 <- factanal(d[,c(-1, -7)], 2)

tab <- cbind(unclass(loadings(fa1)), unclass(loadings(fa2)))
print(
      xtable::xtable(tab, caption = "Factor analysis",
                     label = "tab:fa2", digits = 3),
      booktabs = TRUE,
      file = "tables/fa2.tex"
)
tab
```

### plot

```{r}
tab <- round(cor(d[, -c(1, 7)]), 2)

a <- tril(lv$h_matrix, -1) # strict lower triangular matrix (omit diagonals)
b <- triu(tab, 1) # strict upper triangular matrix
c <- a + b
diag(c) <- NA

pdf("plots/itempairs.pdf", width = 7, height = 7)
plot(as.matrix(c), fmt.cell='%.2f', main = "", key = NULL,
xlab = "Item", ylab = "Item", col = sapply(seq(1, 0, -0.1), gray, alpha = 0.5))
dev.off()

plot(as.matrix(c), fmt.cell='%.2f', main = "", key = NULL,
xlab = "Item", ylab = "Item", col = sapply(seq(1, 0, -0.1), gray, alpha = 0.5))
```

